Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-07-04 ThesisJP/Background/background.tex
Corrected norwegian abstract based on comments from Sverre.

\subsection{Beamforming}
If time delays are applied to the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images that are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$ planes. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

@@

The standard form of beamforming used in ultrasound imaging is known as \textit{delay-and-sum} (DAS): 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\{\Delta_m[n]\}$.  From now on we will omit the dependency on $\{\Delta_m[n]\}$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. When transmitting, the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically have a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations long. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}. Note that after this process the ultrasound data will be complex valued. The total number of samples in a B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challenge with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which support even higher transfer rates, have also recently been introduced.

@@

\begin{figure}[t!]
\subfloat[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfloat[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering echo is coming from 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}

@@

A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3B_wf_c.
\end{align}
For our example, with $L=M/2=64$, this sums up to a required throughput of more than $1/2$ TFLOPS. This is more than thousand times that needed for real-time DAS beamforming. Even though $1/2$ TFLOPS is just 10\% of the peak throughput of modern GPUs, remember that the latter is only a theoretical number, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}. As for DAS beamforming, this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size is typically proportional to the number of array elements.

@@

\subsection{Capon beamforming of focused broadband beams}
Both the LCA and BS-Capon beamformers are based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images. For the LCA beamformer, the windows selected by the Capon beamformer were investigated and a special set of predefined windows were defined. If one looks closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths. The results from applying the LCA beamformer \cite[Fig.\,9]{Synnevag2011} and the success of the BS-Capon beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter. This is mainly due to the focused beams used in medical ultrasound which highly limits the amount of energy arriving off axis. This focusing of energy also explain why Capon beamforming works on broadband ultrasound data.

>>>>

\subsection{Beamforming}
If time delays are applied to the signal emitted by the array elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images that are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$ planes. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

@@

The standard form of beamforming used in ultrasound imaging is known as \textit{delay-and-sum} (DAS): 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a vector of pre-delayed element data, and $z[n]$ is the output image line in the direction given by $\{\Delta_m[n]\}$.  From now on we will omit the dependency on $\{\Delta_m[n]\}$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. When transmitting, the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically have a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations long. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent. Note that after this process the ultrasound data will be complex valued. The total number of samples in a B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challenge with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which support even higher transfer rates, have also recently been introduced.

@@

\begin{figure}[t!]
\subfloat[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfloat[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering echo is coming from 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interferer. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}

@@

A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3B_wf_c.
\end{align}
For our example, with $L=M/2=64$, this sums up to a required throughput of more than $1/2$ TFLOPS, which is more than thousand times that needed for real-time DAS beamforming. On the other hand, $1/2$ TFLOPS is just 10\% of the peak throughput of modern GPUs, but remember that the latter is only a theoretical number, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}. As for DAS beamforming, this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size is typically proportional to the number of array elements.

@@

\subsection{Capon beamforming of focused broadband beams}
Both the LCA and BS-Capon beamformers are based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images. For the LCA beamformer, the windows selected by the Capon beamformer were investigated and a special set of predefined windows were defined. If one looks closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths. The results from applying the LCA beamformer \cite[Fig.\,9]{Synnevag2011} and the success of the BS-Capon beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter. This is mainly due to the focused beams used in medical ultrasound which highly limits the amount of energy arriving off axis. This focusing of energy also explains why Capon beamforming works on broadband ultrasound data.

### 2014-07-04 ThesisJP/Thesis.tex
Corrected norwegian title.

%\thispagestyle{empty}%
%~\\%
%~\\%
%~\\
\noindent\large 
\textbf{Hurtig eksekvering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
%\vspace{0.2cm}\\%
\\\\%
\normalsize
De siste årene har det vært en rivende utvikling innen bruken av skjermkort (GPU-er), som egentlig er laget for å kjøre dataspill, til å gjøre numeriske beregninger for naturvitenskapelige anvendelser. Et sistegenerasjons skjermkort har nå like mye beregningskraft som verdens kraftigste datamaskin hadde rundt år 2000. Utfordringen er at problemet som skal løses må bestå av mange uavhengige deler som kan beregnes i parallell.  For ultralydmaskiner har denne utviklingen gjort at algoritmer som tideligere krevde spesialhardware, nå kan programmeres i software og kjøres i sanntid på et rimelig skjermkort. Et ultralydsystem som bruker skjermkort til å gjøre beregninger vil være mye rimeligere, mer fleksibelt og gjøre det lettere å implementere avansert prosessering enn et system hvor alle algoritmer er og må "hugges i stein".

@@

This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimized GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, accepted for publication}.
	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008}.
	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
\end{enumerate}

@@

%%%% Paper 1 %%%%
\begin{bibunit}[IEEEtran]
\CustChapterPaper{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{IEEE Transactions on Oceanic Engineering, accepted for publication}.
\newpage
{
\input{./Paper1/howto_howto_capon}
}
\newpage\addcontentsline{toc}{section}{References}
\putbib[./Paper1/library]
\end{bibunit}

@@

%%%% Paper 3 %%%%
\begin{bibunit}[IEEEtran]
\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
\runningtitle{Capon Beamforming and Moving Objects}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
\newpage\input{./Paper3/moving_obj}
\newpage\addcontentsline{toc}{section}{References}
\putbib[./Paper3/moving_objbib]
\end{bibunit}

>>>>

%\thispagestyle{empty}%
%~\\%
%~\\%
%~\\
\noindent\large 
\textbf{Hurtig beregning av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
%\vspace{0.2cm}\\%
\\\\%
\normalsize
De siste årene har det vært en rivende utvikling innen bruken av skjermkort (GPU-er), som egentlig er laget for å kjøre dataspill, til å gjøre numeriske beregninger for naturvitenskapelige anvendelser. Et sistegenerasjons skjermkort har nå like mye beregningskraft som verdens kraftigste datamaskin hadde rundt år 2000. Utfordringen er at problemet som skal løses må bestå av mange uavhengige deler som kan beregnes i parallell.  For ultralydmaskiner har denne utviklingen gjort at algoritmer som tideligere krevde spesialhardware, nå kan programmeres i software og kjøres i sanntid på et rimelig skjermkort. Et ultralydsystem som bruker skjermkort til å gjøre beregninger vil være mye rimeligere, mer fleksibelt og gjøre det lettere å implementere avansert prosessering enn et system hvor alle algoritmer er og må "hugges i stein".

@@

This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimized GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, Jul 2014}.
	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 7, pp. 1152-1160, Jul 2014}.
	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008}.
	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
\end{enumerate}

@@

%%%% Paper 1 %%%%
\begin{bibunit}[IEEEtran]
\CustChapterPaper{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{IEEE Transactions on Oceanic Engineering, Jun 2014}.
\newpage
{
\input{./Paper1/howto_howto_capon}
}
\newpage\addcontentsline{toc}{section}{References}
\putbib[./Paper1/library]
\end{bibunit}

@@

%%%% Paper 3 %%%%
\begin{bibunit}[IEEEtran]
\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
\runningtitle{Capon Beamforming and Moving Objects}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, , vol. 61, no. 7, pp. 1152-1160, Jul 2014}.
\newpage\input{./Paper3/moving_obj}
\newpage\addcontentsline{toc}{section}{References}
\putbib[./Paper3/moving_objbib]
\end{bibunit}
